{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-03-30 paper.tex
Datnicek reflections

\subsubsection*{Two-Stage Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for developing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
prototypes at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an additional
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
offer an appealing alternative design.

@@

\subsection{Requirements}
The design of Datnicek brings together a range of recent research ideas on interactive programming
environments for data science. Datnicek notebooks consist of code cells and markdown cells, but
they also support interactive grid cells where the user can use programming by demonstration to
construct data cleaning scripts. We aim to support the following features:

@@

@@

\subsection{Reflections}

edits and formula in interactive should be unified

foof

(data science environment)

what are the limitations of Datnicek

x

x

x

x

@@

NICE THINGS IN DATNICEK
* merging of code edits (e.g. change parameter merges with adding a call to chain)
* remove cell to which there are refs - does the susbtrate catch this?

@@

\section{Discussion}

>>>>

\subsubsection*{Two-Phase Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for developing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second phase to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

@@

\subsubsection*{Concrete Programming}
Programming can be simplified by working with concrete values instead of abstractions, an idea
pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
prototypes at the object level. At the expression level, similar functionality can be provided
by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
mechanism as central, whereas other systems view tracking of copy \& paste as an additional
editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
also tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
for The Gamma \cite{petricek-2020-live}.

@@

\subsection{Requirements}
\label{sec:case-req}

The design of Datnicek brings together a range of recent research ideas on interactive programming
environments for data science. Datnicek notebooks consist of code cells and markdown cells, but
they also support interactive grid cells where the user can use programming by demonstration to
construct data cleaning scripts. We aim to support the following features:

@@

\subsection{Reflections}
The development of Datnicek only started after the design and implementation of Denicek and
Webnicek was completed. Thanks to this two-phase methodology, the experience reported here
provides a qualitative evaluation of the Denicek substrate. It also highlights limitations and
suggests directions for future work.

\subsubsection*{Effectiveness and Unifromity.}
Many of the Datnicek requirements are addressed by Denicek with minimal implementation
effort. The implementation represents all non-transient actions on a notebook as Denicek edits.
Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
changing a primitive value is an edit, and recommendations in the interactive grid are also edits
(appending a {\small\Verb_<transform>_} node).

The uniform and dynamic nature makes the development experience using Denicek more akin to dynamic
object-oriented languages than to statically-typed functional languages. The system can quickly evolve,
but it needs suitable debugging tools, such as the source view (Fig.~\ref{fig:source}) or an
ability to step through the history.

\subsubsection*{Notebook Editing.}
Collaborative editing based on merging of edits makes conflicts less common when compared to
textual merging. For example, changing a method parameter can be merged with adding a call
to a chain. However, adding two independent calls still results in a conflict. A formula
representation that avoids nesting, as in Histogram \cite{petricek-2019-histogram}, may be more flexible.

Variables are represented as references and automatic reference updating %(\S\ref{sec:system-defs})
prevents a number of errors. References are automatically updated when the field name changes.
Denicek also prevents the deletion of cells that contain variables to which there are references.

The proof of concept nature of Datnicek means that it currently lacks suitable user interface for
resolving conflicts. Similarly, the structure editor for code is pointer-based and would benefit from
a better support for keyboard based input \cite{moon-2022-tylr,beckman-2023-sandblocks}.
However, Denicek provides a suitable infrastructure for implementing this.

@@

\subsubsection*{Evaluation Model}
The evaluation model of Denicek implements dependency tracking and incremental recomputation,
which aid notebook reproducibility \cite{petricek-2018-wrattler,koop-2017-dataflow}. As in
Webnicek (\S\ref{sec:impl-incremental}), Datnicek keeps evaluated edits on top of the current history
and use conflict detection to remove invalidated edits. We also uses the same logic for edits
performed in interactive grid cells, suggesting that the approach is a useful general
implementation pattern.

Because invalidation of evaluated edits is not built-in into Denicek, it is worth exploring
if Denicek could support the standard evaluation model of Jupyter notebooks based on mutable
state.

The development of Datnicek uncovered some limitations of the underlying conflict detection
in Denicek. In particular, edits generated by evaluation (such as \ident{Add} edits that set the
evaluated result) need to be treated as conflicting with edits that have both value and
structure effects. This is because they are mere results of a more complex evaluation process.

Data loaded in Datnicek is represented as Denicek document nodes (a table is a list of records).
Such uniform representation makes Denicek more general, but places a high demand on the system
performance. We implemented a number of Denicek optimisations, but Datnicek is still cumbersome
to use with data consisting of multiple thousands of rows. Improving system performance is
thus important future work.

\subsubsection*{Future Work}
One of the design choices discussed in \S\ref{sec:discuss} is whether to track the document
structure explicitly. The structure of documents created in Webnicek can be irregular,
which justified the dynamic nature of Denicek. In contrast, Datnicek notebooks are more regular.
Replicating Datnicek using a system akin to Denicek, but based on explicit structure would
yield an interesting comparison.

Our experience with conflict detection shows the importance of getting the details of the
implementation of programming susbstrates like Denicek right. This could be aided by the
development of a formally tractable model of the system and proving a correctness property
akin to that of The Gamma live previews \cite{petricek-2020-foundations}.

Finally, the representation of interactive grid transformations (Fig.~\ref{fig:source})
highlights a duplication -- a transformation is represented as a sequence of edits,
but also as a formula. An interesting future direction is unifying or integrating those two
kinds of computation in the Denicek substrate.

@@

\subsection{Olsen}
Situations, Tasks, Users + Importance

Generality - webnicek, denicek, map space via TDPSs
Flexibility - dynamic nature via edits - experienced in Datnicek
Expressive Match - possibly - for document based things
Can it Scale up - remains a question

\subsection{Technical Dimensions}
TDPS determine a space of designs
how much of this space can be covered by systems based on Denicek?

INTERACTION
Feedback Loops
Modes of Interaction
Abstraction Construction

NOTATION
Notational Structure

Surface and Internal Notation
Uniformity of Notations

CONCEPTUAL STRUCTRUE
not much to say
Conceptual integrity vs openness - integrity
Composability Convenience

CUSTOMIZABILITY
self-sustainability - future work

COMPLEXITY
Factoring of Complexity - no reuse for now, remains to be explored

\section{Conclusion}

knowledge exists only in the brain of system designers!!

@@
